An individual differences perspective on the development of pragmatic abilities in the preschool years

### 2022-04-07 paper/manuscript.Rmd
prep submission

authornote: |
  We are very thankful to Stella Christie for sharing the material for the relational match-to-sample task with us.

  M. Bohn received funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Sklodowska-Curie grant agreement no. 749229. M. H. Tessler was funded by the National Science Foundation SBE Postdoctoral Research Fellowship Grant No. 1911790. M. C. Frank was supported by a Jacobs Foundation Advanced Research Fellowship and the Zhou Fund for Language and Cognition. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

abstract: |
  Pragmatic abilities are fundamental to successful language use and learning. Individual differences studies contribute to understanding the psychological processes involved in pragmatic reasoning as well as the link to other cognitive abilities. Small sample sizes, insufficient measurement tools, and a lack of theoretical precision have hindered progress. Three studies addressed these challenges in a sample of three- to five-year-old German-speaking children (*N* = 228, 121 female). Studies 1 and 2 assessed the psychometric properties of six pragmatics tasks. Study 3 investigated relations among pragmatics tasks and between pragmatics and other cognitive abilities. The shared cognitive processes were formalized in a computational cognitive model of pragmatic reasoning. Results showed substantive relations among pragmatic tasks and between pragmatic abilities and executive functions.

keywords          : "Pragmatics, language development, individual differences, cognitive modeling"
wordcount         : "X"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

@@

```{r}
data <- bind_rows(
  read_csv("../data/data_r1.csv") %>%filter(task != "training") %>%mutate(study = "study1"),
  read_csv("../data/data_r2.csv") %>%filter(task != "training")%>%mutate(study = "study2"),
  read_csv("../data/data_r3.csv") %>%filter(task != "training")%>%mutate(study = "study3")
  ) %>%
  mutate(study = factor(study),
         id = paste(study, id, sep = "_"), 
         test_day = paste0("day", test_day))
```
# Introduction

Communication predates language. Before children produce their first words, they communicate with the world around them using vocalizations and gestures [@bruner1974communication; @bates1979emergence]. The process of language learning recruits many of the social-cognitive processes that underlay pre-verbal communication [@bohn2019pervasive; @tomasello2009constructing; @clark2009first]. Even for proficient language users is communication not reducible to the words being exchanged. The common thread running through these different aspects of human communication is its inferential nature: what a speaker means -- verbal or otherwise -- is underdetermined by the parts that make up the utterance. It takes contextual social inferences to recover the intended meaning [@grice1991studies; @clark1996using; @sperber2001relevance; @levinson2000presumptive; @bohn2018common]. These inferences are often referred to as *pragmatic* inferences [e.g., @bohn2019pervasive].

The development of pragmatics has been widely studied in recent years [for a recent review see @bohn2019pervasive]. This research covers a range of different phenomena ranging from so-called *pure pragmatics* [@matthews2014pragmatic] in non-verbal communication in infancy to sophisticated linguistic inferences developing much later [@huang2009online; @papafragou2016scalar]. A growing portion of this work is devoted to studying individual differences [@matthews2018individual; @wilson2021pragmatic]. The motivation behind this is twofold: on the one hand, individual differences offer insights into the underlying psychological processes [@kidd2018individual; @matthews2018individual; @wilson2022novel]. On the other hand, deficits in pragmatic abilities have been linked to behavioral problems and forms of language impairment [@helland2014stable].

In their recent review, @matthews2018individual identified three issues that significantly limit what we can learn from individual differences research on pragmatic abilities. First, most studies have insufficient sample sizes so that small and medium sized correlations among pragmatics tasks and between pragmatics tasks and measures for other cognitive abilities cannot be reliably detected. Second, the tasks used to assess pragmatic abilities often have poor or unknown psychometric properties. For example, many tasks only have a single trial and are therefore unable to capture variation between children. Reliability is not assessed, making it unclear if the task captures stable characteristics [see also @russell2008assessing]. Third, the cognitive processes underlying pragmatic inferences in a particular task are underspecified. As a consequence, there is no clear rationale for why a task should correlate with another cognitive measure.

The studies presented here directly address the issues identified by Matthews and colleagues [@matthews2018individual]. We identified six pragmatic reasoning tasks in children between three and five years of age and investigated their psychometric properties, in particular their re-test reliability. Reliable tasks are a necessary precondition for meaningful individual differences research [@hedge2018reliability; @fried2018measurement]. Next, we investigated the relations among different pragmatic reasoning tasks as well as between pragmatic reasoning and executive functions in a sample large enough to detect small to medium sized correlations. Finally, we introduced computational cognitive models of pragmatic reasoning to the study of individual differences. This allowed us to formalize the cognitive processes we think underlay pragmatic reasoning and thereby provide a substantive theoretical account for why certain pragmatic reasoning tasks are related to one another and why executive functions relate to them.

The six tasks we selected were developmental adaptations of referential communication games inspired by research in experimental pragmatics [@noveck2008experimental; @noveck2004experimental]. They all share a common trial-by-trial structure in which the test event always involved an agent producing an ambiguous utterance that the child had to resolve using pragmatic reasoning. This allowed us to run multiple trials per task, and with that, to capture individual differences. We grouped the tasks into two broad categories (Figure \@ref(fig:fig1)). On the one hand, utterance-based tasks asked children to derive inferences from the words and gestures the speaker produced in context. On the other hand, common ground/discourse-based tasks asked children to derive inferences from the social interaction that preceded the utterance.

For the utterance-based category, we selected a mutual exclusivity, informativeness inference, and ad-hoc implicature task. Mutual exclusivity -- or principle of contrast -- describes the phenomenon that children tend to map a novel word to an unknown object [@markman1988children; @halberda2003development; @bion2013fast; @merriman1989mutual; @clark1988logic; @lewis2020role]. Informativeness inferences describe situations in which children identify the referent of a novel word by assuming that the speaker is trying to be informative. Being informative translates to using words that have smaller extensions in context [@frank2014inferring]. Ad-hoc implicature describes inferences that ask the child to contrast an utterance with alternatives that the speaker could have used but did not [@stiller2015ad; @yoon2019role; @katsos2011pragmatic].

For the discourse-based category, we selected a speaker preference, discourse novelty and discourse continuity task. In the speaker preference task, the child had to track the preference of a speaker in order to identify the referent of a novel word [@saylor2009preschoolers]. Discourse novelty refers to a situation in which the child tracks the temporal appearance of objects and expects the speaker to refer to objects that are new in context [@akhtar1996role; @diesendruck2004two]. In the discourse continuity task, the child had to infer and track the topic of an ongoing conversation to resolve ambiguity [@bohn_le_peloquin_koymen_frank_2020; @akhtar2002relevance].

In addition to the pragmatics tasks, we also included tasks for executive functions [@zelazo2006dimensional] and of analogical reasoning [@christie2014language]. Executive functions refer to a family of top-down mental processes that enable us to inhibit automatic or intuitive responses and allow us to concentrate and focus attention on particulars [@diamond2013executive]. A substantial body of research has investigated the link between executive functions and pragmatics -- with mixed results [@matthews2018individual; @nilsen2009relations]. Analogical reasoning refers to the ability to reason about abstract relations between stimuli [@carstensen2021graded] -- an ability that has not been linked to pragmatics -- at least not to the same extent as executive functions.

After finding in Study 1 and 2 that most of the pragmatics tasks showed good to excellent re-test reliability, we studied a larger sample of children in Study 3 to investigate relations between the three utterance-based tasks. We focused on these tasks for theoretical reasons: we assume that -- computationally -- they share a common contrastive inference process according to which the listener compares the utterance the speaker made to alternative possible utterances. The listener expects the speaker to be helpful and informative, that is, to choose the utterance that best communicates the intended message. We formalize these assumptions in a computational cognitive model which we then use to study individual differences in this alleged process. Study 3 also included tasks for executive functions and analogical reasoning. Across analytical approaches, we found systematic relations among the pragmatics tasks as well as between pragmatics and executive functions, but not analogical reasoning.

@@

Study 1 focused on the psychometric properties of four pragmatics tasks, in particular, their re-test reliability. two of the tasks were from the utterance-based group and two from the common ground/discourse-based group. As a fifth task, we included a measure of executive functions. Methods and sample size were pre-registered at https://osf.io/6a723. All analysis scripts and data files can be found in the following repository: https://github.com/manuelbohn/pragBat. The same repository also contains the code to run the experiments.

@@

For Study 1, we collected data from `r dem%>%filter(study == "study1")%>%pull(n)` children ($m_{age}$ = `r dem%>%filter(study == "study1")%>%pull(mage)`, range$_{age}$: `r dem%>%filter(study == "study1")%>%pull(lrange)` - `r dem%>%filter(study == "study1")%>%pull(urange)`, `r dem%>%filter(study == "study1")%>%pull(female)` girls) of which `r dem%>%filter(study == "study1")%>%pull(reli_n)` were tested twice. For most children, the two test sessions were two days apart; the longest time difference was six days. Children came from an ethnically homogeneous, mid-size German city (~550,000 inhabitants, median income €1,974 per month as of 2020); were mostly monolingual and had mixed socioeconomic backgrounds. The study was approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology. Data was collected between November 2019 and January 2020.

@@

(ref:figlab1) Overview of the tasks used in Study 1 to 3. Pictures show screenshots from each task. The vertical order corresponds to the order of presentation in each study. The colors group the tasks along the (assumed) cognitive processes involved. Blue: utterance-based inferences. Red common ground/discourse-based inferences. Green: Executive functions. Yellow: Analogical reasoning.

@@

The task was adapted from @bohn_tessler_merrick_frank_2019. Th animal was standing between two trees with objects hanging in them. In one tree, there were two objects (type A and B) and in the other tree there was only one (type B). The animal turned to the tree with the two objects and labelled one of the objects. It was unclear from the animal's utterance, which of the two objects they were referring to. We assumed that children would map the novel word onto the object of type A because they expected the animal to turn to the tree with only the object of type B if their intention was to provide a label for an object of type B. Next, the trees were replaced by new ones, one of which carried an object of type A and the other of type B. The animal then said that one of the trees had the same object as they labelled previously (using the same label) and asked the child to touch the tree. We coded as correct if the child selected the tree with the object of type A. The first two trials were training trials, in which there was only one object in each tree. There were five test trials. The location of the tree with the two objects in the beginning of each trial was pseudo-randomized and so was the location of the objects when the new trees appeared.

@@

This task was modeled after @zelazo2006dimensional. The child saw to cards, a blue rabbit on the left and a red boat on the right. The experimenter introduced the child to the color game they would be playing next. In this game, all blue cards (irrespective of object depicted) would go to the left card and all red cards to the right. Next, a third card appeared in the middle of the screen (red rabbit or blue boat) and the experimenter demonstrated the color sorting by moving the card to the one with the same color. After a second demonstration trial, the child started to do the color sorting by themselves. After six trials, the experimenter said that they were now going to play a different game, the shape game, according to which all rabbits would go to the card with the rabbit (left) and all boats to the card with the boat (right). The experimenter repeated these instructions once and without any demonstration the child continued with the sorting according to the new rule. There were six test trials. The shape on the card was pseudo-randomized across trials. We only coded the trials after the rule change and coded as correct when the child sorted according to shape.

@@

We analysed the data in three steps. First we investigated developmental effects in each task, then we assessed re-test reliability and finally, we looked at relations between the tasks. All analysis were run in `R` [@R-base] version 4.1.2. Regression models were fit as Bayesian generalized linear mixed models (GLMM) using the function `brm` from the package `brms` [@burkner2017brms]. We used default priors for all analysis.

To estimate developmental effects in each task, we fit a GLMM predicting correct responses (0/1) by age (centered at the mean) and trial number (also centered). The model included random intercepts for each participant and random slopes for trial within participants (model notation in `R`: `correct ~ age + trial + (trial|id)` )^[We pre-registered the inclusion of random intercepts for item. We deviate from this here because the order of items was fixed and the same for all participants so that trial and item were confounded.]. For each task, we inspected and visualized the posterior distribution (mean and 95% Credible Interval (CrI)) for the age estimate.

We assessed re-test reliability in two ways. First, for each task we computed the proportion of correct trials for each individual in the two test sessions and then used  Pearson correlations to quantify re-test reliability. Second, we used a GLMM based approach suggested by @rouder2019psychometrics. Here, a GLMM was fitted to the trial-by-trial data for each task with a fixed effect of age, a random intercept for each participant and a random slope for test day (`correct ~ age + (0+test_day|id)`)^[The notation `0+test_day` yields a separate intercept estimate for each test day and subject instead of an intercept estimate for day 1 and a slope for the difference between day 1 and day 2. As a consequence, the model estimates the correlation between the two test days instead of a correlation between an intercept and the slope for test day.]. The model yields a participant specific estimate for each test day and also estimates the correlation between the two. This correlation can be interpreted as the re-test reliability. This approach has several advantages. First, it uses the trial-by-trial data and avoids information loss that comes with data aggregation. Second, it uses hierarchical shrinkage to obtain better participant specific estimates. Finally, it allows us to get an age-independent estimate for reliability. One worry when assessing re-test reliability in developmental studies is that re-test correlations can be high because of domain general cognitive gains and not because of task-specific individual differences. By including age as a fixed effect in the model, the estimates for each participant are independent of age and so is the correlation between estimates for the two test days -- the re-test reliability.

Finally, we used that aggregated data from both test days for each participant and task to compute Pearson correlations between the different tasks. Given the small sample size in Study 1, this part of the analysis was mostly exploratory.

@@

(ref:figlab2) Results by task for studies 1 to 3. Each panel shows the results for one task. Regression lines show the predicted developmental trajectories (with 95% CrI) based on by-task GLMMs, with the line type indicating the study. Colored points show age group means (with 95% CI based on non-parametric bootstrap) with the different shapes corresponding to the different studies. Light shapes show the mean performance for each subject by study. Dotted line shows level of performance expected by chance.

@@

We found developmental effects in most of the tasks. Figure \@ref(fig:fig2) shows the data and visualizes the developmental trajectories based on the model. Figure \@ref(fig:fig3) shows the model estimates for age. In the mutual exclusivity task, performance was reliably above chance level and increased with age. For informative inference, the pattern was quite different: Performance was at chance level with only minor developmental gains. In the speaker preference task, performance was again clearly above chance with developmental gains resulting in a ceiling effect for older children. In the discourse novelty task, performance was also above chance with no clear developmental effects. The card sorting task showed the strongest developmental effects with younger children performing largely below chance and older children performing above chance.

Re-test reliability was high for most tasks (see Figure \@ref(fig:fig4)). Raw correlation between the two test sessions was above .7 for mutual exclusivity, speaker preference and discourse novelty. With .62 it was slightly lower for card sorting. The model based -- age independent -- reliability estimates yielded similar results suggesting that the tasks did capture task specific individual differences. A notable exception was the informativeness inference task, which was not reliable according to any of the methods of computing re-test reliability (Figure \@ref(fig:fig4)). We suspected the overall low variation in performance to be responsible for this.

@@

Most correlations between the tasks were low and ranged between *r* = -0.2 and 0.2 (see Figure \@ref(fig:fig2)). A notable exception was the correlation between mutual exclusivity and card sorting with *r* = `r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(rho)` (95% CI[`r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(lci)` - `r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(uci)`]).

(ref:figlab3) Model estimates (with 95% CrI) for age based on GLMMs for each task and study.

@@

Study 1 showed that the different tasks were age appropriate and reliable. A notable exception was the informativeness inference task which generated no systematic variation in the age range we studied here. Correlations between the tasks were generally low, with the notable exception of the relation between mutual exclusivity and card sorting. Given the small sample size, we want to avoid overly strong claims, however, it was interesting to see that the relation between the two tasks tapping into discourse-based inferences (speaker preference and discourse novelty) were -- if anything -- negatively correlated.

@@

Based on these results od Study 1, we compiled a new set of tasks for Study 2. We retained the mutual exclusivity and card sorting tasks because of the interesting relation between the two found in Study 2. We simplified the informativeness inference task to be more age appropriate with the hope to induce more variation in performance. We removed the speaker preference and discourse novelty tasks -- despite their excellent re-test reliability -- because they seemed to be unrelated to one another and also unrelated to the other tasks. The new tasks focused on ad-hoc implicature and discourse continuity. As noted in the introduction, we had theoretical reasons to expected the ad-hoc implicature task to be related to the mutual exclusivity and informativeness inference tasks. We had no such strong predictions for the discourse continuity task.

@@

Participants for Study 2 were recruited from the same general population. We collected data from `r dem%>%filter(study == "study2")%>%pull(n)` children ($m_{age}$ = `r dem%>%filter(study == "study2")%>%pull(mage)`, range$_{age}$: `r dem%>%filter(study == "study2")%>%pull(lrange)` - `r dem%>%filter(study == "study2")%>%pull(urange)`, `r dem%>%filter(study == "study2")%>%pull(female)` girls) of which `r dem%>%filter(study == "study2")%>%pull(reli_n)` were tested twice. The two test sessions were again two days apart; the longest time difference was 14 days. Data was collected between March and October 2020.

@@

We used the same methods to analyse the data as in Study 1.

@@

From Study 2, we used the the mutual exclusivity, ad-hoc implicature, informativeness inference and card sorting tasks. We added the relational match-to-sample task, which we now describe in more detail.

>>>>

authornote: |
  We are very thankful to Stella Christie for sharing the material for the relational match-to-sample task with us. M. Bohn received funding from the European Union’s Horizon 2020 research and innovation programme under the Marie Sklodowska-Curie grant agreement no. 749229. M. H. Tessler was funded by the National Science Foundation SBE Postdoctoral Research Fellowship Grant No. 1911790. M. C. Frank was supported by a Jacobs Foundation Advanced Research Fellowship and the Zhou Fund for Language and Cognition. The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

abstract: |
  Pragmatic abilities are fundamental to successful language use and learning. Individual differences studies contribute to understanding the psychological processes involved in pragmatic reasoning as well as the link to other cognitive abilities. Small sample sizes, insufficient measurement tools, and a lack of theoretical precision have hindered progress, however. Three studies addressed these challenges in a sample of three- to five-year-old German-speaking children (*N* = 228, 121 female). Studies 1 and 2 assessed the psychometric properties of six pragmatics tasks. Study 3 investigated relations among pragmatics tasks and between pragmatics and other cognitive abilities. The shared cognitive processes were formalized in a computational cognitive model of pragmatic reasoning. The tasks were found to measure stable variation between individuals on a shared latent construct relating to pragmatic inference.

keywords          : "Pragmatics, language development, individual differences, cognitive modeling"
wordcount         : "8491"

@@

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_docx
---

@@

```{r}
data <- bind_rows(
  read_csv("../data/data_r1.csv") %>%filter(task != "training") %>%mutate(study = "study1"),
  read_csv("../data/data_r2.csv") %>%filter(task != "training")%>%mutate(study = "study2"),
  read_csv("../data/data_r3.csv") %>%filter(task != "training")%>%mutate(study = "study3")
  ) %>%
  mutate(study = factor(study),
         id = paste(study, id, sep = "_"), 
         test_day =  factor(test_day))
```
# Introduction

Communication predates language. Before children produce their first words, they communicate with the world around them using vocalizations and gestures [@bruner1974communication; @bates1979emergence]. The process of language learning recruits many of the social-cognitive processes that underlie pre-verbal communication [@bohn2019pervasive; @tomasello2009constructing; @clark2009first]. Even for proficient language users, communication not reducible to the words being exchanged. The common thread running through these different aspects of human communication is its inferential nature: what a speaker means -- verbal or otherwise -- is underdetermined by the parts that make up the utterance. It takes contextual social inferences to recover the intended meaning [@grice1991studies; @clark1996using; @sperber2001relevance; @levinson2000presumptive; @bohn2018common]. These inferences are often referred to as *pragmatic* inferences.

The development of pragmatics has been widely studied in recent years [for a recent review see @bohn2019pervasive]. This research covers a range of different phenomena ranging from so-called *pure pragmatics* [@matthews2014pragmatic] in non-verbal communication in infancy to sophisticated linguistic inferences developing much later [@huang2009online; @papafragou2016scalar]. A growing portion of this work is devoted to studying individual differences [@matthews2018individual; @wilson2021pragmatic]. The motivation behind the move to study individual variation is twofold: on the one hand, individual differences offer insights into the underlying psychological processes. If two phenomena (e.g. pragmatic reasoning and executive functions) vary together this is consistent with shared cognitive processes [@kidd2018individual; @matthews2018individual; @wilson2022novel], though it is not definitive evidence for such a claim. On the other hand, deficits in pragmatic abilities have been linked to behavioral problems and forms of language impairment [@helland2014stable].

In their recent review, @matthews2018individual identified three issues that significantly limit what we can learn from individual differences research on pragmatic abilities. First, most studies have insufficient sample sizes so that small and medium sized correlations among pragmatics tasks and between pragmatics tasks and measures for other cognitive abilities cannot be reliably detected [mirroring issues in estimating correlations across other fields, @schonbrodt2013sample]. Second, the tasks used to assess pragmatic abilities often have poor or unknown psychometric properties. For example, many tasks only have a single trial and are therefore unable to capture variation between children [see also @enkavi2019large]. Reliability is not assessed, making it unclear if the task captures stable characteristics [@russell2008assessing; @flake2020measurement]. Third, the cognitive processes underlying pragmatic inferences in a particular task are underspecified. As a consequence, there is often no clear rationale for why a particular target task should correlate with another cognitive measure.

In search of a better understanding of individual variation in pragmatic ability, the studies presented here directly address these issues. We identified six pragmatic reasoning tasks in children between three and five years of age and investigated their psychometric properties, in particular their re-test reliability. Reliable tasks are a necessary precondition for meaningful individual differences research [@hedge2018reliability; @fried2018measurement]. Next, we investigated the relations among different pragmatic reasoning tasks as well as between pragmatic reasoning and other cognitive abilities in a sample large enough to detect small to medium sized correlations.

 Finally, we introduced computational cognitive models of pragmatic reasoning to the study of individual differences. Computational cognitive models formalize hypotheses about cognitive processes that could underlie pragmatic reasoning; thus, the use of these models provides a substantive theoretical account for why certain pragmatic reasoning tasks should be related to one another. Here, we use the formalism introduced by the Rational Speech Act (RSA) framework [@frank2012predicting; @goodman2016pragmatic]. RSA models see pragmatic inferences as a special case of (Bayesian) social reasoning. A pragmatic listener interprets an utterance by assuming it was produced by a cooperative speaker. The speaker tries to be informative, that is, they provide messages that would increase the probability that the listener will recover their intended meaning. The informativeness of an utterance arises from a contrastive inference in which the effects of multiple – plausible – utterances are compared. We assume that this inference process is shared by some of the pragmatics tasks involved in this study and can thus be used to account for individual differences (see below).

The six tasks we selected were developmental adaptations of referential communication games inspired by research in experimental pragmatics [@noveck2008experimental; @noveck2004experimental]. They all share a common trial-by-trial structure in which the test event always involved an agent producing an ambiguous utterance that the child had to resolve using pragmatic reasoning. This structure allowed us to run multiple trials per task, increasing reliability. We grouped the tasks into two broad categories (Figure \@ref(fig:fig1)). On the one hand, *utterance-based tasks* asked children to derive inferences from the words and gestures the speaker produced in context. On the other hand, *common ground/discourse-based tasks* asked children to derive inferences from the social interaction that preceded the utterance.

For the utterance-based category, we selected mutual exclusivity, informativeness inference, and ad-hoc implicature tasks. "Mutual exclusivity"^[Following @lewis2020role, we use the term "mutual exclusivity" as a convenient term to denote a specific task. This term is also related to a particular theoretical account of the phenomenon [@markman1990constraints], but we do not presuppose that specific account.] describes the phenomenon that children tend to map a novel word to an unknown object [@markman1988children; @halberda2003development; @bion2013fast; @merriman1989mutual; @clark1988logic; @lewis2020role]. Informativeness inferences describe situations in which children identify the referent of a novel word by assuming that the speaker is trying to be informative. Being informative translates to using words that have smaller extensions in context [@frank2014inferring]. Ad-hoc implicature describes inferences that ask the child to contrast an utterance with alternatives that the speaker could have used but did not [@stiller2015ad; @yoon2019role; @katsos2011pragmatic].

For the discourse-based category, we selected speaker preference, discourse novelty and discourse continuity tasks. In the speaker preference task, the child had to track the preference of a speaker in order to identify the referent of a novel word [@saylor2009preschoolers]. Discourse novelty refers to a situation in which the child tracks the temporal appearance of objects and expects the speaker to refer to objects that are new in context [@akhtar1996role; @diesendruck2004two]. In the discourse continuity task, the child had to infer and track the topic of an ongoing conversation to resolve ambiguity [@bohn_le_peloquin_koymen_frank_2020; @akhtar2002relevance].

In addition to the pragmatics tasks, we also included two additional cognitive tasks: one measuring executive functions [@zelazo2006dimensional] and the other analogical reasoning [@christie2014language]. Executive functions refer to a family of top-down mental processes that enable us to inhibit automatic or intuitive responses and allow us to concentrate and focus attention on particulars [@diamond2013executive]. A substantial body of research has investigated the link between executive functions and pragmatics -- with mixed results [@matthews2018individual; @nilsen2009relations]. Analogical reasoning refers to the ability to reason about abstract relations between stimuli [@carstensen2021graded] -- an ability that, to our knowledge, has not been specifically linked to pragmatics -- at least not to the same extent as executive functions.

After finding in Study 1 and 2 that most of the pragmatics tasks showed relatively good re-test reliability, we studied a larger sample of children in Study 3 to investigate relations between the three utterance-based tasks. We focused on these tasks for theoretical reasons: as noted above, we assume that -- computationally -- they share a common contrastive inference process. We formalize these assumptions in a computational cognitive model which we then use to study individual differences in this alleged process. Study 3 also included tasks for executive functions and analogical reasoning. Across analytical approaches, we found systematic relations among the pragmatics tasks as well as between pragmatics and executive functions, but not analogical reasoning. In the discussion, we use the structure of the cognitive model to speculate about the psychological processes shared between pragmatics and executive functions.

@@

Study 1 focused on the psychometric properties of four pragmatics tasks, in particular, their re-test reliability. Two of the tasks were from the utterance-based group and two from the common ground/discourse-based group. As a fifth task, we included a measure of executive functions. Methods and sample size were pre-registered at https://osf.io/6a723. All analysis scripts and data files can be found in the following repository: https://github.com/manuelbohn/pragBat. The same repository also contains the code to run the experiments.

@@

For Study 1, we collected data from `r dem%>%filter(study == "study1")%>%pull(n)` children ($m_{age}$ = `r dem%>%filter(study == "study1")%>%pull(mage)`, range$_{age}$: `r dem%>%filter(study == "study1")%>%pull(lrange)` - `r dem%>%filter(study == "study1")%>%pull(urange)`, `r dem%>%filter(study == "study1")%>%pull(female)` girls), of whom `r dem%>%filter(study == "study1")%>%pull(reli_n)` were tested twice. For most children, the two test sessions were two days apart; the longest time difference was six days. Children came from an ethnically homogeneous, mid-size German city (~550,000 inhabitants, median income €1,974 per month as of 2020); were mostly monolingual and had mixed socioeconomic backgrounds. The study was approved by an internal ethics committee at the Max Planck Institute for Evolutionary Anthropology. Data was collected between November 2019 and January 2020.

@@

(ref:figlab1) Overview of the tasks used in Study 1 to 3. Pictures show screenshots from each task. The vertical order corresponds to the order of presentation in each study. The colors group the tasks along the (assumed) cognitive processes involved. Blue: utterance-based inferences. Red common ground/discourse-based inferences. Green: Executive functions. Yellow: Analogical reasoning. Bold numbers show the number of trials per task.

@@

The task was adapted from @bohn_tessler_merrick_frank_2019. The animal was standing between two trees with objects hanging in them. In one tree, there were two objects (type A and B) and in the other tree there was only one (type B). The animal turned to the tree with the two objects and labeled one of the objects. It was unclear from the animal's utterance, which of the two objects they were referring to. We assumed that children would map the novel word onto the object of type A because they expected the animal to turn to the tree with only the object of type B if their intention was to provide a label for an object of type B. Next, the trees were replaced by new ones, one of which carried an object of type A and the other of type B. The animal then said that one of the trees had the same object as they labeled previously (using the same label) and asked the child to touch the tree. We coded as correct if the child selected the tree with the object of type A. The first two trials were training trials, in which there was only one object in each tree. There were five test trials. The location of the tree with the two objects in the beginning of each trial was pseudo-randomized and so was the location of the objects when the new trees appeared.

@@

This task was modeled after @zelazo2006dimensional. The child saw two cards, a blue rabbit on the left and a red boat on the right. The experimenter introduced the child to the color game they would be playing next. In this game, all blue cards (irrespective of objects depicted) would go to the left card and all red cards to the right. Next, a third card appeared in the middle of the screen (red rabbit or blue boat) and the experimenter demonstrated the color sorting by moving the card to the one with the same color. After a second demonstration trial, the child started to do the color sorting by themselves. After six trials, the experimenter said that they were now going to play a different game, the shape game, according to which all rabbits would go to the card with the rabbit (left) and all boats to the card with the boat (right). The experimenter repeated these instructions once and without any demonstration the child continued with the sorting according to the new rule. There were six test trials. The shape on the card was pseudo-randomized across trials. We only coded the trials after the rule change and coded as correct when the child sorted according to shape.

@@

We analyzed the data in three steps. First we investigated developmental effects in each task, then we assessed re-test reliability, and finally, we looked at relations between the tasks. All analyses were run in `R` [@R-base] version 4.1.2. Regression models were fit as Bayesian generalized linear mixed models (GLMM) using the function `brm` from the package `brms` [@burkner2017brms]. We used default priors for all analysis.

To estimate developmental effects in each task, we fit a GLMM predicting correct responses (0/1) by age (in years, centered at the mean) and trial number (also centered). The model included random intercepts for each participant and random slopes for trial within participants (model notation in `R`: `correct ~ age + trial + (trial|id)` )^[We pre-registered the inclusion of random intercepts for item. We deviate from this here because the order of items was fixed and the same for all participants so that trial and item were confounded.]. For each task, we inspected and visualized the posterior distribution (mean and 95% Credible Interval (CrI)) for the age estimate.

We assessed re-test reliability in two ways. First, for each task we computed the proportion of correct trials for each individual in the two test sessions and then used Pearson correlations to quantify re-test reliability. Second, we used a GLMM based approach suggested by @rouder2019psychometrics. Here, a GLMM was fitted to the trial-by-trial data for each task with a fixed effect of age (in years, centered at the mean), a random intercept for each participant and a random slope for test day (`correct ~ age + (0+test_day|id)`)^[The notation `0+test_day` yields a separate intercept estimate for each test day and subject instead of an intercept estimate for day 1 and a slope for the difference between day 1 and day 2. As a consequence, the model estimates the correlation between the two test days instead of a correlation between an intercept and the slope for test day.]. The model yields a participant specific estimate for each test day and also estimates the correlation between the two. This correlation can be interpreted as the re-test reliability. This approach has several advantages. First, it uses trial-by-trial data and avoids information loss that comes with data aggregation. Second, it uses hierarchical shrinkage to obtain better participant-specific estimates. Finally, it allows us to get an age-independent estimate for reliability. One worry when assessing re-test reliability in developmental studies is that re-test correlations can be high because of domain general cognitive gains and not because of task-specific individual differences. By including age as a fixed effect in the model, the estimates for each participant are independent of age and so is the correlation between estimates for the two test days -- the re-test reliability.

Finally, we used aggregated data from both test days for each participant and task to compute Pearson correlations between the different tasks. Given the small sample size in Study 1, this part of the analysis was mostly exploratory.

@@

(ref:figlab2) Results by task for studies 1 to 3. Each panel shows the results for one task. Regression lines show the predicted developmental trajectories (with 95% CrI) based on by-task GLMMs, with the line type indicating the study. Colored points show age group means (with 95% CI based on non-parametric bootstrap) with the different shapes corresponding to the different studies. Light shapes show the mean performance for each subject by study. Dotted line shows the level of performance expected by chance.

@@

We found developmental effects in most of the tasks. Figure \@ref(fig:fig2) shows the data and visualizes the developmental trajectories based on the model. Figure \@ref(fig:fig3) shows the model estimates for age. In the mutual exclusivity task, performance was reliably above chance level and increased with age. For informativeness inference, the pattern was quite different: Performance was at chance level with only minor developmental gains. In the speaker preference task, performance was again clearly above chance with developmental gains resulting in a ceiling effect for older children. In the discourse novelty task, performance was also above chance with no clear developmental effects. The card sorting task showed the strongest developmental effects with younger children performing largely below chance and older children performing above chance.

Re-test reliability was high for most tasks (see Figure \@ref(fig:fig4)). Raw correlation between the two test sessions was above .7 for mutual exclusivity, speaker preference and discourse novelty, though it was slightly lower for card sorting (.62). The model based -- age independent -- reliability estimates yielded similar results suggesting that the tasks did capture task specific individual differences. A notable exception was the informativeness inference task, which was not reliable according to any of the methods of computing re-test reliability (Figure \@ref(fig:fig4)). We suspected the overall low variation in performance to be responsible for this.

@@

Most correlations between the tasks were low and ranged between *r* = -0.2 and 0.2 (see Figure \@ref(fig:fig2)). A notable exception was the correlation between mutual exclusivity and card sorting (*r* = `r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(rho)`, 95% CI[`r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(lci)` - `r cors%>%filter(study == "study1", task1 == "mutual_exclusivity", task2 == "card_sorting")%>%pull(uci)`]).

(ref:figlab3) Model estimates (with 95% CrI) for age (in years, centered at the mean) based on GLMMs for each task and study.

@@

Study 1 showed that the different tasks were – for the most part – age appropriate and reliable. A notable exception was the informativeness inference task which generated no systematic variation in the age range we studied here. Correlations between the tasks were generally low, with the exception of the relation between mutual exclusivity and card sorting. Given the small sample size, we avoid overly strong claims, however, it was interesting to see that the relation between the two tasks tapping into discourse-based inferences (speaker preference and discourse novelty) were -- if anything -- negatively correlated.

@@

Based on these results of Study 1, we compiled a new set of tasks for Study 2. We retained the mutual exclusivity and card sorting tasks because of the interesting relation between the two found in Study 2. We simplified the informativeness inference task to be more age appropriate with the hope of inducing more variation in performance. We removed the speaker preference and discourse novelty tasks -- despite their excellent re-test reliability -- because they seemed to be unrelated to one another and also unrelated to the other tasks. We also added new tasks focused on ad-hoc implicature and discourse continuity. As noted in the introduction, we had theoretical reasons to expect the ad-hoc implicature task to be related to the mutual exclusivity and informativeness inference tasks. We had no such strong predictions for the discourse continuity task.

@@

Participants for Study 2 were recruited from the same general population. We collected data from `r dem%>%filter(study == "study2")%>%pull(n)` children ($m_{age}$ = `r dem%>%filter(study == "study2")%>%pull(mage)`, range$_{age}$: `r dem%>%filter(study == "study2")%>%pull(lrange)` - `r dem%>%filter(study == "study2")%>%pull(urange)`, `r dem%>%filter(study == "study2")%>%pull(female)` girls), of whom `r dem%>%filter(study == "study2")%>%pull(reli_n)` were tested twice. The two test sessions were again two days apart; the longest time difference was 14 days. Data was collected between March and October 2020.

@@

We used the same methods to analyze the data as in Study 1.

@@

From Study 2, we used the mutual exclusivity, ad-hoc implicature, informativeness inference and card sorting tasks. We added the relational match-to-sample task, which we now describe in more detail.
